Parsing Postfix log files

### 2009-03-12 doc/logparser-architecture.tex
Corrections to the actions section of the implementation chapter. Removed a large chunk explaining postfix_action; remaining references will need to be excised over time. Renamed some labels, resulting in some churn without any real effect; in particular some long paragraphs were reformatted.

Separating the rules from the actions and framework makes it possible to
parse new inputs without modifying the core parsing algorithm.  Adding a
new rule with the action to invoke and a regex to recognise the inputs is
trivial in comparison to understanding an entire parser, identifying the
correct location to change, and making the appropriate changes.  Additions
to a parser must be made without adversely affecting existing parsing,
including any edge cases that are not immediately obvious; an edge case
that occurs only four times in \numberOFlogFILES{} log files is described
in Yet More Aborted Delivery Attempts
(\sectionref{yet-more-aborted-delivery-attempts}).  More intrusive changes
are more likely to introduce a bug, so reducing the extent of the changes
is important.  Requiring changes to a parser's source code also complicates
upgrades of the parser, because the changes must be preserved during the
upgrade, and they may clash with changes made by the developer.  This
architecture allows the user to add new rules without having to edit a
parser, unless the new inputs cannot be processed by the existing actions.
If the new inputs do require new functionality, new actions can be added to
the parser without having to modify existing actions; only in the rare
event that the new actions need to cooperate with existing actions will
more extensive changes be required.

@@

During development of \parsername{} it became apparent that in addition to
the obvious variety in log~lines there were many complications to be
overcome.  Some were the result of deficiencies in Postfix's logging, and
some of those deficiencies were rectified by later versions of Postfix,
e.g.\ identifying bounce notifications
(\sectionref{identifying-bounce-notifications}); others were due to the
vagaries of process scheduling, client behaviour, and administrative
actions.  All were successfully accommodated in \parsername{}: adding new
actions was enough to overcome several of the complications; others
required modifications to a single existing action to work around a
difficulty; the remainder were resolved by adapting existing actions to
cooperate and exchange extra data (via the framework), changing their
behaviour as appropriate based on that extra data.  Every architecture
should aim to make the easy things easy and the hard things possible; the
successful implementation of \parsername{} demonstrates that this
architecture achieves that aim.

>>>>

Separating the rules from the actions and framework makes it possible to
parse new inputs without modifying the core parsing algorithm.  Adding a
new rule with the action to invoke and a regex to recognise the inputs is
trivial in comparison to understanding an entire parser, identifying the
correct location to change, and making the appropriate changes.  Additions
to a parser must be made without adversely affecting existing parsing,
including any edge cases that are not immediately obvious; an edge case
that occurs only four times in \numberOFlogFILES{} log files is described
in Yet More Aborted Delivery Attempts
(\sectionref{yet more aborted delivery attempts}).  More intrusive changes
are more likely to introduce a bug, so reducing the extent of the changes
is important.  Requiring changes to a parser's source code also complicates
upgrades of the parser, because the changes must be preserved during the
upgrade, and they may clash with changes made by the developer.  This
architecture allows the user to add new rules without having to edit a
parser, unless the new inputs cannot be processed by the existing actions.
If the new inputs do require new functionality, new actions can be added to
the parser without having to modify existing actions; only in the rare
event that the new actions need to cooperate with existing actions will
more extensive changes be required.

@@

During development of \parsername{} it became apparent that in addition to
the obvious variety in log~lines there were many complications to be
overcome.  Some were the result of deficiencies in Postfix's logging, and
some of those deficiencies were rectified by later versions of Postfix,
e.g.\ identifying bounce notifications (\sectionref{identifying bounce
notifications}); others were due to the vagaries of process scheduling,
client behaviour, and administrative actions.  All were successfully
accommodated in \parsername{}: adding new actions was enough to overcome
several of the complications; others required modifications to a single
existing action to work around a difficulty; the remainder were resolved by
adapting existing actions to cooperate and exchange extra data (via the
framework), changing their behaviour as appropriate based on that extra
data.  Every architecture should aim to make the easy things easy and the
hard things possible; the successful implementation of \parsername{}
demonstrates that this architecture achieves that aim.

### 2009-03-12 doc/logparser-implementation.tex
Corrections to the actions section of the implementation chapter. Removed a large chunk explaining postfix_action; remaining references will need to be excised over time. Renamed some labels, resulting in some churn without any real effect; in particular some long paragraphs were reformatted.

    \item Postfix logs enough information to make it possible to accurately
        reconstruct the actions it has taken.  There are several heuristics
        used when parsing; see
        \sectionref{identifying-bounce-notifications},
        \sectionref{aborted-delivery-attempts}, and \sectionref{pickup
        logging after cleanup} for details.

@@

\label{flow-chart}

@@

\label{mail-enters-the-system}

@@

Internally generated mails lack any explicit origin in Postfix 2.2.x and
must be detected using heuristics as described in
\sectionref{identifying-bounce-notifications}.  Bounce notifications are
the primary example of internally generated mails, though there may be
other types.  Postfix may generate mails to the administrator when it
encounters configuration errors, but such mails are presumably rare.

@@

    \item If one or more mails were accepted there will be more log lines
        for those mails later, see \sectionref{mail-delivery}.

@@

\label{mail-delivery}

@@

    \item [postfix\_action] The action Postfix must have taken to generate
        this log line; the possible actions are described later in
        \sectionref{list of postfix actions in implementation}.

@@

The actions reconstruct the journey a mail takes through Postfix.  Details
of the actions available in the Postfix parser can be found in
\sectionref{actions in detail in implementation}, and \sectionref{adding
new actions in implementation} describes the process of adding new actions.

In the Postfix log parser developed for this project there are
\numberOFactions{} actions and \numberOFrules{} rules, with an uneven
distribution of rules to actions as shown in \graphref{Distribution of
rules per action}.  Unsurprisingly, the action with the most associated
rules is \texttt{DELIVERY\_REJECTED}, the action that handles Postfix
rejecting a mail delivery attempt; it is followed by \texttt{SAVE\_DATA},
the action responsible for handling informative log lines, supplementing
the data gathered from other log lines.  The third most common action is,
perhaps surprisingly, \texttt{UNINTERESTING}: this action does nothing when
executed, allowing uninteresting log lines to be parsed without causing any
effects (it does not imply that the input is ungrammatical or unparsed).
Generally rules specifying the \texttt{UNINTERESTING} action parse log
lines that are not associated with a specific mail, e.g.\ notices about
configuration files changing.  The remaining actions have only one or two
associated rules: some actions are required to address a deficiency in the
log files, or a complication that arises during parsing;  other actions
will only ever have one log line variant, e.g.\ all log lines showing that
a remote client has connected are matched by a single rule and handled by
the \texttt{CONNECT} action.

Using the \texttt{CONNECT} action as an example: it creates a new data
structure in memory for the new client connection, saving the data
extracted by the rule into it; this data will be entered into the database
when the mail delivery attempt is complete.  If a data structure already
exists for the new connection it is treated as a symptom of a bug, and the
action issues a warning containing the full contents of the existing data
structure, plus the log line that has just been parsed.

Each action is passed the same arguments:

@@

    \item [rule] The matching rule.

    \item [matches] The data extracted from the line by the rule's regex.

@@

        \begin{eqlist}[\def\makelabel#1{\bfseries#1:}]

@@

            \item [host] The hostname of the server that logged the line.

@@

            \item [text] The remainder of the line.

        \end{eqlist}

@@

Each action saves all data extracted by the rule's regex, if there is
enough information in the log line to identify the correct connection to
save the data to.

@@

    \item [BOUNCE] Postfix 2.3 and subsequent versions log the creation of
        bounce messages.  This action creates a new mail if necessary; if
        the mail already exists the unknown origin flag will be removed.
        The action also marks the mail as a bounce notification.  To deal
        with complication \sectionref{Bounce notification mails delivered
        before their creation is logged} this action checks a cache of
        recent bounce mails to avoid creating bogus bounce mails when log
        lines are out of order.

@@

    \item [COMMIT] Enter the data from the mail into the database.  Entry
        will be postponed if the mail is a child waiting to be tracked.
        Once entered, the mail will be deleted from the state tables.
        Deletion will be postponed if the mail is the parent of re-injected
        mail (\sectionref{Re-injected mails}).

    \item [CONNECT] Handle a remote client connecting: create a new state
        table entry (indexed by \daemon{smtpd} \acronym{pid}) and save both
        the client hostname and \acronym{IP} address.

    \item [DELETE] Deals with mail deleted using Postfix's administrative
        command \daemon{postsuper}.  This action adds a dummy recipient
        address if required, then invokes the COMMIT action to handle
        adding the mail to the database.  The complication this action
        deals with is described fully in \sectionref{Mail deleted before
        delivery is attempted}.

    \item [DELIVERY\_REJECTED] Deals with Postfix rejecting an
        \acronym{SMTP} command from the remote client: log the rejection
        with a mail if there is a queueid in the log line, or with the
        connection if not.

    \item [DISCONNECT] Deal with the remote client disconnecting: enter the
        connection in the database, perform any required cleanup, and
        delete the connection from the state tables.  This action deals
        with aborted delivery attempts
        (\sectionref{aborted-delivery-attempts}).

    \item [EXPIRY] If Postfix has not managed to deliver a mail after
        trying for five days it will give up and return the mail to the
        sender.  When this happens the mail will not have a combination of
        Postfix programs which passes the valid combinations check (see
        \sectionref{out of order log lines}).  To ensure that the mail can
        be committed the EXPIRY action sets a flag marking the mail as
        expired; the flag later causes the valid combinations check to be
        skipped, so the mail will be committed.

    \item [MAIL\_QUEUED] This action represents Postfix picking a mail from
        the queue to deliver.  This action is used for both \daemon{qmgr}
        and \daemon{cleanup} as it needs to deal with out of order log
        lines; see \sectionref{discarding cleanup log lines} for details.
        XXX EXPLAIN THIS BETTER\@.

    \item [MAIL\_TOO\_LARGE] When a client tries to send a mail larger than
        the local server accepts it will be discarded and the client
        informed of the problem.  This is handled in exactly the same way
        as the TIMEOUT action, further details are given there.

    \item [PICKUP] The PICKUP action corresponds to the \daemon{pickup}
        service dealing with a locally submitted mail.  Out of order log
        entries may have caused the state table entry to already exist (see
        \sectionref{pickup logging after cleanup}); otherwise it is
        created.  The data extracted from the log line is then saved to the
        state table entry.

    \item [POSTFIX\_RELOAD] When Postfix stops or reloads its configuration
        it kills all \daemon{smtpd} processes,\footnote{Possibly other
        programs are killed also, but the parser is only affected by and
        interested in \daemon{smtpd} processes exiting.} requiring any
        active connections to be cleaned up, entered in the database, and
        deleted from the state tables.

    \item [SAVE\_DATA] Find the correct mail based on the queueid in
        the log line, and save the data extracted by the regex to it.

    \item [SMTPD\_DIED] Sometimes a \daemon{smtpd} dies or exits
        unsuccessfully; the active connection for that \daemon{smtpd} must
        be cleaned up, entered in the database, and deleted from the state
        tables.

    \item [SMTPD\_KILLED] Sometimes an \daemon{smtpd} is killed by a
        signal; the active connection for
        that \daemon{smtpd} must be cleaned up, entered in the database,
        and deleted from the state tables.

    \item [SMTPD\_WATCHDOG] \daemon{smtpd} processes have a watchdog timer
        to deal with unusual situations --- after five hours the timer will
        expire and the \daemon{smtpd} will exit.  This occurs very
        infrequently, as there are many other timeouts that should occur
        in the intervening hours: DNS timeouts, timeouts reading data
        from the client, etc\@.  The active connection for that
        \daemon{smtpd} must be cleaned up, entered in the database, and
        deleted from the state tables.

    \item [TIMEOUT] The connection timed out so the mail currently being
        transferred must be discarded.  The mail may have been accepted, in
        which case there's a data structure to dispose of, or it may not in
        which case there is not.  See
        \sectionref{timeouts-during-data-phase} for the gory details.

    \item [TRACK] Track a mail when it is re-injected for forwarding to
        another mail server; this happens when a local address is aliased
        to a remote address.  TRACK will be called when dealing with the
        parent mail, and will create the child mail if necessary.  TRACK
        checks if the child has already been tracked, either by this parent
        or by another parent, and issues appropriate warnings in either
        case.

@@

The distribution of rules per action is shown in \graphref{Distribution of
rules per action}.

\subsection{Postfix Actions}

\label{list of postfix actions in implementation}

XXX WHY DO I NEED HAVE postfix\_actions AT ALL\@?

XXX SEE TABLE BELOW\@: THERE IS NEARLY A 1--1 MAPPING BETWEEN action AND
postfix\_action.  NEW ACTIONS COULD BE ADDED TO FORCE A 1--1 MAPPING, E.G.
COMMIT + DISCARDED COULD BECOME A DISCARD ACTION\@.  MAIL\_QUEUED HAVING
INFO AND PROCESSING CAN PROBABLY BE DEALT WITH, THE BIGGEST DIFFICULTY WILL
BE WITH SAVE\_DATA\@.

\begin{tabular}{lll}

1 & BOUNCE & BOUNCED \\
1 & CLONE & ACCEPTED \\
3 & COMMIT & DISCARDED \\
1 & COMMIT & INFO \\
1 & CONNECT & INFO \\
1 & DELETE & DELETED \\
64 & DELIVERY\_REJECTED & REJECTED \\
1 & DISCONNECT & INFO \\
1 & EXPIRY & EXPIRED \\
1 & MAIL\_QUEUED & INFO \\
1 & MAIL\_QUEUED & PROCESSING \\
1 & MAIL\_TOO\_LARGE & DISCARDED \\
1 & PICKUP & INFO \\
2 & POSTFIX\_RELOAD & POSTFIX\_RELOAD \\
13 & SAVE\_DATA & BOUNCED \\
38 & SAVE\_DATA & INFO \\
1 & SAVE\_DATA & REJECTED \\
5 & SAVE\_DATA & SENT \\
2 & SMTPD\_DIED & IGNORED \\
1 & SMTPD\_WATCHDOG & IGNORED \\
2 & TIMEOUT & DISCARDED \\
1 & TRACK & SENT \\
42 & UNINTERESTING & IGNORED \\

postfix\_action IS USED IN A FEW ACTIONS, BUT IT COULD PROBABLY BE EASILY
REMOVED, OR MAYBE I SHOULD KEEP TRACK OF actions INSTEAD\@?

XXX IF I KEEP THESE I NEED TO EXPLAIN WHY I HAVE THEM\@.

    \item [ACCEPTED] Postfix has accepted a mail, and will
        subsequently attempt to deliver it.

    \item [BOUNCED] The mail has bounced, because of a mail loop,
        delivery failure, or five day timeout.\footnote{As with
        much of Postfix's behaviour, five days is the default value
        but can be changed by the administrator if they choose.}

    \item [DELETED] The mail was deleted from the queue by an
        administrator.

    \item [DISCARDED] Postfix discarded the mail it was in the
        process of accepting, because it was either larger than the
        size limit set by the administrator, or the client timed
        out or disconnected.

    \item [EXPIRED] The mail has been in the queue for five days without
        successful delivery.  A bounce mail will be generated and sent to
        the sender address.

    \item [INFO] Represents an unspecified intermediate action that
        the parser is not interested in per se, but that does log
        useful information, supplementing other log lines.

    \item [IGNORED] An action that is not only uninteresting in
        itself, but that also provides no useful data.

    \item [POSTFIX\_RELOAD] The administrator has instructed
        Postfix to start or stop, and all \daemon{smtpd} processes
        will be terminated.  This does not negatively affect the
        log files or mail queued by Postfix for delivery.

    \item [PROCESSING] \daemon{cleanup} is processing a mail.
        XXX ADD MORE DETAIL HERE

    \item [REJECTED] Postfix rejected a command from the remote
        client, causing at least one recipient to be rejected.

    \item [SENT] Postfix has successfully sent a mail.

@@

Adding new actions is not as easy as adding new rules, though care has been
taken in the parser's design and implementation to make adding new actions
as painless as possible.  The implementer writes a subroutine that accepts
the standard arguments given to actions, and registers it as an action by
calling the framework subroutine add\_actions() with the name of the new
action as a parameter.  The new action must be registered before the rules
are loaded, because it is an error for a rule to specify an unregistered
action; this helps catch mistakes made when adding new rules.  No other
work is required from the implementer to integrate the action into the
parser; all of their attention and effort can be focused on correctly
implementing their action.  The only negative aspect is that the process
involves editing the parser source code, which makes upgrading to a later
version of the parser more difficult, though by no means impossible.  If
the author of the new action wishes, they can take advantage of the
parser's object oriented implementation by sub-classing it and implementing
their changes in the derived class, allowing future upgrades of the parser
with greatly reduced chance of conflicts.\footnote{The real difference
between the two approaches is where the new code is placed.  The simpler
option is to change the parser code directly, but those changes will then
have to be made to subsequent versions of the parser, and as the scope of
the changes increases so does the chance of conflict, or mistakes when
copying the action.  The more time consuming option is to write a subclass
containing the new actions and change the program that invokes the parser
so that it uses the subclass rather than the parser; the changes required
to the program invoking the parser are minor and much less likely to lead
to conflicts when upgrading to future versions of the parser.  An
alternative is to submit new actions to the author of the parser for
inclusion in future versions, resulting in two benefits: the new actions do
not need to be maintained separately, and other users of the parser can
avail of the new functionality.} The action may need to extend the list of
valid combinations described in \sectionref{out of order log lines} if the
new action creates a different set of acceptable programs, but this is
unlikely to occur, as it would require parsing log lines from Postfix
components the parser currently ignores.\footnote{The mail server used for
development does not utilise either the \daemon{lmtp} or \daemon{virtual}
delivery agents, so this parser does not have rules to handle log lines
from those components.  Adding new rules to parse those component's log
lines is a simple process, though if their behaviour differs significantly
from the \daemon{smtp} or \daemon{local} delivery agents new actions may be
required.  The mail server in question is a production mail server handling
mail for a university department; the benefit of using this server is that
the log files used exhibit the idiosyncrasies and peculiarities a mail
server in the wild must deal with, but the downside is that significantly
altering the configuration just to log messages from a different Postfix
component is not an option.}

@@

Multiple independent mails may be delivered across one connection: this
requires the algorithm to clone the current data as soon as a mail is
accepted, so that subsequent mails will not trample over each other's data.
This must be done every time a mail is accepted, as it is impossible to
tell in advance which connections will accept multiple mails.  Once the
mail has been accepted its log lines will not be correlated by
\acronym{pid} any more, its queueid will be used instead (except when
timeouts occur during the data phase
\sectionref{timeouts-during-data-phase}).  If the original connection has
any useful data (e.g.\ rejections) it will be saved to the database when
the client disconnects.  One unsolved difficulty is distinguishing between
different groups of rejections, e.g.\ when dealing with the following
sequence:

@@

    \item The indicator which more commonly introduces re-injection is when
        \daemon{qmgr} selects a mail with a previously unseen queueid for
        delivery, in which case a new data structure will be created for
        that mail.  The mail will be flagged as having unknown origins;
        this flag will subsequently be cleared once the origin has been
        established.  This may also be an indicator that the mail is a
        bounce notification, see
        \sectionref{identifying-bounce-notifications} for details.

@@

\label{identifying-bounce-notifications}

@@

\label{aborted-delivery-attempts}

@@

Some mail clients disconnect abruptly if a second or subsequent recipient
is rejected; they may also disconnect after other errors, but such
disconnections are either unimportant or are handled elsewhere in the
parser (\sectionref{timeouts-during-data-phase}).  Sadly, Postfix does not
log a message saying the mail has been discarded, as should be expected by
now.  The checks to identify this happening are:

@@

\label{timeouts-during-data-phase}

@@

The circumstances above account for all occasions identified thus far where
a \daemon{smtpd} suddenly stops logging.  In addition to removing an active
connection the last accepted mail may need to be discarded, as detailed in
\sectionref{timeouts-during-data-phase}; otherwise the queueid state table
is untouched.

@@

\label{yet-more-aborted-delivery-attempts}

The aborted delivery attempts described in
\sectionref{aborted-delivery-attempts} occur frequently, but the aborted
delivery attempts described in this section only occur four times in the
\numberOFlogFILES{} log files used for testing.  The symptoms are the same
as in \sectionref{aborted-delivery-attempts}, except that there
\textit{is\/} a \daemon{cleanup} log line; there does not appear to be
anything in the log file to explain why there are no further log lines.
The only way to detect these mails is to periodically scan all mails in the
state tables, deleting any mails displaying the following characteristics:

@@

    \item Improve the heuristics used in
        \sectionref{timeouts-during-data-phase}, or develop another
        solution, to avoid incorrectly leaving a mail in the state tables.

>>>>

    \item Postfix logs enough information to make it possible to accurately
        reconstruct the actions it has taken.  There are several heuristics
        used when parsing; see \sectionref{identifying bounce
        notifications}, \sectionref{aborted delivery attempts}, and
        \sectionref{pickup logging after cleanup} for details.

@@

\label{flow chart}

@@

\label{mail enters the system}

@@

Internally generated mails lack any explicit origin in Postfix 2.2.x and
must be detected using heuristics as described in \sectionref{identifying
bounce notifications}.  Bounce notifications are the primary example of
internally generated mails, though there may be other types.  Postfix may
generate mails to the administrator when it encounters configuration
errors, but such mails are presumably rare.

@@

    \item If one or more mails were accepted there will be more log lines
        for those mails later, see \sectionref{mail delivery}.

@@

\label{mail delivery}

@@

@@

Actions are the component of this architecture responsible for processing
all of the inputs recognised by rules; in \parsername{} they reconstruct
the journey each mail takes through Postfix, dealing with all the
complications and difficulties that arise.  \parsername{} has
\numberOFactions{} actions and \numberOFrules{} rules, with an uneven
distribution of rules to actions as shown in \graphref{Distribution of
rules per action}.  Unsurprisingly, the action with the most associated
rules is \texttt{DELIVERY\_REJECTED}, the action that processes Postfix
rejecting a mail delivery attempt; next is \texttt{SAVE\_DATA}, the action
responsible for processing informative log lines, supplementing the data
gathered from other log lines.  The third most common action is, perhaps
surprisingly, \texttt{UNINTERESTING}: this action does nothing when
invoked, allowing uninteresting log lines to be parsed without any effects
(it does not imply that the input is ungrammatical or unparsed).  Generally
rules specifying the \texttt{UNINTERESTING} action recognise log lines that
are not associated with a specific mail, e.g.\ notices about configuration
files changing; these log lines are recognised and processed so that the
framework can warn about unrecognised log lines, informing the user that
they need to add to or extend their ruleset.  The remaining actions have
only one or two associated rules: some are specialised actions required to
address a deficiency in the log files, or a complication that arises during
parsing; other actions will only ever have one log line variant, e.g.\ all
log lines showing that a remote client has connected are recognised by a
single rule and handled by the \texttt{CONNECT} action.  \parsernames{}
actions are described in detail in \sectionref{actions in detail in
implementation}, and \sectionref{adding new actions in implementation}
covers the process of adding a new action.

This section explains the actions found in \parsername{}; it may help to
revisit the flow chart in \sectionref{flow chart} to see how a mail passes
from one action to another as its log lines are recognised.  The words
\textit{mail\/} and \textit{connection\/} are used in the actions
descriptions below because they are less unwieldy and more specific than
\textit{state table entry\/}; a connection becomes a mail during the
\texttt{CLONE} action, which processes Postfix accepting a delivery
attempt, and the data structure moves from one state table to another
within the parser.

The complications and difficulties that arose when parsing log files
generated on a production mail server are documented in
\sectionref{complications}; some action descriptions refer to specific
difficulties they address.  The complications are documented in a separate
section to avoid overwhelming the action descriptions.

If there is enough information in the log line to identify the correct
connection or mail, each action will save all the data captured by the
recognising rule's regex.  Each action is passed the same arguments:

@@

    \item [rule] The recognising rule.

    \item [data] The data captured from the log line by the rule's regex.

@@

@@

@@

            \item [host] The hostname of the server the line was logged on.

            \item [text] The remainder of the log line, i.e.\ the message
                logged by the program and recognised by the rule.

@@

@@

    \item [BOUNCE\_CREATED] Postfix 2.3 and subsequent versions log the
        creation of bounce messages.  This action creates a new mail if
        necessary; if the mail already exists the unknown origin flag will
        be removed.  The action also marks the mail as a bounce
        notification.  To deal with complication \sectionref{Bounce
        notification mails delivered before their creation is logged} this
        action checks a cache of recent bounce mails, to avoid creating
        bogus bounce mails when log lines are out of order.

@@

    \item [COMMIT] Enter the data from the mail into the database.  Entry
        will be postponed if the mail is a child waiting to be tracked
        (\sectionref{Re-injected mails}).  Once entered in the database,
        the mail will be usually be deleted from the state tables, but
        deletion will be postponed if the mail is the parent of re-injected
        mail (\sectionref{Re-injected mails}).

    \item [CONNECT] Handle a remote client connecting: create a new
        connection indexed by \daemon{smtpd} \acronym{pid}, and save the
        data captured by the rule (\acronym{IP} address and hostname of the
        client).  If a connection already exists it is treated as a symptom
        of a bug, and the action issues a warning containing the full
        contents of the existing connection plus the log line that has
        just been parsed.

    \item [DELETE] Deals with mail deleted using Postfix's administrative
        command \daemon{postsuper}.  This action adds a dummy recipient
        address if required, then invokes the COMMIT action to save the
        mail to the database.  The complication this action deals with is
        described fully in \sectionref{Mail deleted before delivery is
        attempted}.

    \item [DELIVERY\_REJECTED] Deals with Postfix rejecting an
        \acronym{SMTP} command from the remote client: log the rejection
        with a mail if there is a queueid in the log line, or with the
        connection if not.  XXX VERY IMPORTANT ACTION\@; EXPAND AND
        REWRITE\@.

    \item [DISCONNECT] Invoked when the remote client disconnects, it
        enters the connection in the database, performs any required
        cleanup, and deletes the connection from the state tables.  This
        action deals with aborted delivery attempts (\sectionref{aborted
        delivery attempts}).

    \item [EXPIRY] If Postfix has not managed to deliver a mail after
        trying for five days it will give up and return the mail to the
        sender.  When this happens the mail will not have a combination of
        Postfix programs which passes the valid combinations check (see
        \sectionref{out of order log lines}).  To ensure that the mail can
        be committed the EXPIRY action sets a flag marking the mail as
        expired; this flag later causes the valid combinations check to be
        skipped.  The mail will be committed when a log line triggers the
        \texttt{COMMIT} action; if the \texttt{COMMIT} action has already
        run and refused to enter the mail in the database because it failed
        the valid program combinations check, the \texttt{SAVE\_DATA}
        action will re-invoke the \texttt{COMMIT} action the next time it
        is invoked.

    \item [MAIL\_BOUNCED] XXX WRITE THIS\@; INVOKES SAVE\_DATA ACTION\@.

    \item [MAIL\_DISCARDED] XXX WRITE THIS\@; INVOKES COMMIT ACTION\@.

    \item [MAIL\_QUEUED] This action represents Postfix picking a mail from
        the queue to deliver.  This action is used for both \daemon{qmgr}
        and \daemon{cleanup} because it needs to deal with out of order log
        lines; see \sectionref{discarding cleanup log lines} for details.
        XXX EXPLAIN THIS BETTER\@.

    \item [MAIL\_SENT] XXX WRITE THIS\@; INVOKES SAVE\_DATA ACTION\@.

    \item [MAIL\_TOO\_LARGE] When a client tries to send a mail larger than
        the local server accepts it will be discarded and the client
        informed of the problem.  The mail may have been accepted and
        partially transferred, in which case there's a data structure to
        dispose of, otherwise there will not be.  See \sectionref{timeouts
        during data phase} for the gory details; although that describes
        timeouts the processing is the same for mails that are too large.

    \item [PICKUP] The PICKUP action corresponds to the \daemon{pickup}
        service dealing with a locally submitted mail.  Out of order log
        entries may have caused the state table entry to already exist (see
        \sectionref{pickup logging after cleanup}); if it does not exist it
        will be created.

    \item [POSTFIX\_RELOAD] When Postfix stops or reloads its configuration
        it kills all \daemon{smtpd} processes, requiring any active
        connections to be cleaned up, entered in the database, and deleted
        from the state tables.  Postfix probably kills all the other
        components too, but the parser is only affected by \daemon{smtpd}
        processes exiting.

    \item [SAVE\_DATA] Every action that can locate the correct entry in
        the state tables saves any data captured by the recognising rule's
        regex to it.  The \texttt{SAVE\_DATA} saves data in this way but
        does not do anything else; it is invoked for log lines that contain
        useful data but do not require any further processing.

    \item [SMTPD\_DIED] Sometimes a \daemon{smtpd} dies or exits
        unsuccessfully; the active connection for that \daemon{smtpd} must
        be cleaned up, entered in the database if it has sufficient data,
        and deleted from the state tables.  In some cases the connection
        will not have enough data to satisfy the database schema, so it
        cannot be entered into the database; unfortunately this means that
        some data is lost.

    \item [SMTPD\_KILLED] Sometimes an \daemon{smtpd} is killed by a
        signal; this is processed in the same way as the
        \texttt{SMTPD\_DIED} action.

    \item [SMTPD\_WATCHDOG] \daemon{smtpd} processes have a watchdog timer
        to deal with unusual situations; after five hours the timer will
        expire and the \daemon{smtpd} will exit.  This occurs very
        infrequently, as there are many other timeouts that should occur in
        the intervening hours, e.g.\ timeouts for DNS requests, timeouts
        reading data from the client.  The active connection for that
        \daemon{smtpd} must be cleaned up, entered in the database, and
        deleted from the state tables.

    \item [TIMEOUT] The connection timed out so the mail currently being
        transferred must be discarded.  The mail may have been accepted, in
        which case there's a data structure to dispose of, or it may not in
        which case there is not.  See \sectionref{timeouts during data
        phase} for the gory details.

    \item [TRACK] Track a mail when it is re-injected for forwarding to
        another mail server; this happens when a local address is aliased
        to a remote address (see \sectionref{tracking re-injected mail} for
        a full explanation).  \texttt{TRACK} will be called when dealing
        with the parent mail, and will create the child mail if necessary.
        \texttt{TRACK} checks if the child has already been tracked, either
        by this parent or by another parent, and issues appropriate
        warnings in either case.

@@

@@

Adding new actions is not as easy as adding new rules, though care has been
taken in the architecture and implementation to make adding new actions as
painless as possible.  The implementer writes a subroutine that accepts the
standard arguments given to actions, and registers it with the framework.
The new action must be registered before the rules are loaded, because it
is an error for a rule to specify an unregistered action; this helps catch
mistakes made when adding new rules.  No other work is required from the
implementer to integrate the action into the parser; all of their attention
and effort can be focused on correctly implementing their action.  The only
negative aspect is that the process involves editing the parser source
code, which makes upgrading to a later version of the parser more
difficult, though by no means impossible.  If the author of the new action
wishes, they can take advantage of the parser's object oriented
implementation by sub-classing it and implementing their changes in the
derived class, allowing future upgrades of the parser with greatly reduced
chance of conflicts.\footnote{The real difference between the two
approaches is where the new code is placed.  The simpler option is to
change the parser code directly, but those changes will then have to be
made to subsequent versions of the parser, and as the scope of the changes
increases so does the chance of conflict or mistakes.  The second option
requires more effort initially but less effort in the long run: write a
subclass containing the new actions and change the program that invokes the
parser so that it uses the subclass rather than the parent class; the
changes required are minor and much less likely to lead to conflicts when
upgrading to future versions of the parser.  An alternative is to submit
new actions to the author \parsername{} for inclusion in future versions so
that the new actions do not need to be maintained separately, and other
users of the parser can benefit from the new functionality.} The action may
need to extend the list of valid combinations described in \sectionref{out
of order log lines} if the new action creates a different set of acceptable
programs, but this is unlikely to occur, as it would require parsing log
lines from Postfix components the parser currently ignores.\footnote{The
mail server used for development does not utilise either the \daemon{lmtp}
or \daemon{virtual} delivery agents, so this parser does not have rules to
handle log lines from those components.  Adding new rules to parse those
component's log lines is a simple process, though if their behaviour
differs significantly from the \daemon{smtp} or \daemon{local} delivery
agents new actions may be required.  The mail server in question is a
production mail server handling mail for a university department; the
benefit of using this server is that the log files used exhibit the
idiosyncrasies and peculiarities a mail server in the wild must deal with,
but the downside is that significantly altering the configuration just to
log messages from a different Postfix component is not an option.}

@@

Multiple independent mails may be delivered across one connection: this
requires the algorithm to clone the current data as soon as a mail is
accepted, so that subsequent mails will not trample over each other's data.
This must be done every time a mail is accepted, as it is impossible to
tell in advance which connections will accept multiple mails.  Once the
mail has been accepted its log lines will not be correlated by
\acronym{pid} any more, its queueid will be used instead (except when
timeouts occur during the data phase \sectionref{timeouts during data
phase}).  If the original connection has any useful data (e.g.\ rejections)
it will be saved to the database when the client disconnects.  One unsolved
difficulty is distinguishing between different groups of rejections, e.g.\
when dealing with the following sequence:

@@

    \item The indicator which more commonly introduces re-injection is when
        \daemon{qmgr} selects a mail with a previously unseen queueid for
        delivery, in which case a new data structure will be created for
        that mail.  The mail will be flagged as having unknown origins;
        this flag will subsequently be cleared once the origin has been
        established.  This may also be an indicator that the mail is a
        bounce notification, see
        \sectionref{identifying bounce notifications} for details.

@@

\label{identifying bounce notifications}

@@

\label{aborted delivery attempts}

@@

Some mail clients disconnect abruptly if a second or subsequent recipient
is rejected; they may also disconnect after other errors, but such
disconnections are either unimportant or are handled elsewhere in the
parser (\sectionref{timeouts during data phase}).  Sadly, Postfix does not
log a message saying the mail has been discarded, as should be expected by
now.  The checks to identify this happening are:

@@

\label{timeouts during data phase}

@@

The circumstances above account for all occasions identified thus far where
a \daemon{smtpd} suddenly stops logging.  In addition to removing an active
connection the last accepted mail may need to be discarded, as detailed in
\sectionref{timeouts during data phase}; otherwise the queueid state table
is untouched.

@@

\label{yet more aborted delivery attempts}

The aborted delivery attempts described in \sectionref{aborted delivery
attempts} occur frequently, but the aborted delivery attempts described in
this section only occur four times in the \numberOFlogFILES{} log files
used for testing.  The symptoms are the same as in \sectionref{aborted
delivery attempts}, except that there \textit{is\/} a \daemon{cleanup} log
line; there does not appear to be anything in the log file to explain why
there are no further log lines.  The only way to detect these mails is to
periodically scan all mails in the state tables, deleting any mails
displaying the following characteristics:

@@

    \item Improve the heuristics used in \sectionref{timeouts during data
        phase}, or develop another solution, to avoid incorrectly leaving a
        mail in the state tables.
